Cluster Difference Imaging Photometric Survey. III. The Coeval Halo and Core of NGC\,2516

### 2021-01-27 cluster_paper/ms.tex
add words

@@

%% Reintroduced the \received and \accepted commands from AASTeX v5.2.
%% Add "Submitted to " argument.
\received{---}
\revised{---}
\accepted{---}
\submitjournal{AAS journals.}
\shorttitle{Core \& Halo of NGC\,2516}

@@

\begin{abstract}
  foo,bar.
\end{abstract}

@@

In the traditional picture of star formation molecular clouds go
gravitationally unstable, and collapse into little knots.  The knots
produce many stars close to each other.  The resulting ``open star
clusters'' are, as with any first-order guess, spherical.

How does the cluster evolve?  The answer depends
on the total stellar mass.
% What exactly does this mean?
The smallest 90\% of star clusters disperse during the ``embedded
phase'' (CITE).  This is driven mostly by {\it process A} (CITE).  The
more massive clusters make it to maybe 100 Myr. Maybe a bit longer.
Their evaporation is thought to be driven by collisions with molecular
clouds (CITE, Spitzer 1958), and the galactic tide (CITE), and {\it
process A}, and {\it process B}.  Of course, the expected evaporation
time should depend on factors including the mass of the cluster
itself, how many high-mass (O and B) stars form, since their winds
[and maybe supernovae] clear out most of the cloud, and the initial
density of the gas cloud to begin with.

@@

We've recently been making TESS light curves of age-dated stars across
the sky, as part of a Cluster Difference Imaging Photometry Survey
(CDIPS, CITE Bouma+19).  Our analysis of Cycle 1 (Sectors 1-13)
yielded light curves of \numsouthernuniqlcs candidate cluster members
in the Southern Ecliptic hemisphere, available on MAST
[\href{https://archive.stsci.edu/hlsp/cdips}{1},\href{https://ui.adsabs.harvard.edu/abs/2019ApJS..245...13B/abstract}{2}].
% TODO: make this number actually quantitative...
Based on rotation periods, $\approx$25\% appear to be bonafide cluster
members.

As part of a broader project of identifying a large and clean sample
of young stars for a transit search, we take the opportunity of this
paper to answer a rather modest question: in just a single rich
southern open cluster, is the cluster halo coeval with the core?  The
cluster we chose for this analysis was NGC\,2516, since it was young
($\sim$110\,Myr) and close ($d=400$\,pc) enough to facilitate rotation
measurements using TESS, and some degree of spectroscopic analysis.
We want to know: is the halo real? To what extent can we use Gaia
alone to reliably identify age-dated needles in the haystack of boring
field stars?  And more generally, what are the implications for the
evolution of open clusters if they do have halos?

Section~\ref{sec:gaia} presents the astrometric and photometric data
from Gaia, and clarifies our usage of the terms ``core'' and ``halo''.
Section~\ref{sec:agedate} age-dates the halo of NGC\,2516, using TESS
gyrochronology (Section~\ref{sec:rotation}), and lithium depletion
(Section~\ref{sec:lithium}).  In Section~\ref{sec:discussion} we
discuss the implications of this analysis for NGC\,2516 specifically
and stellar spin-down and open cluster evolution generally.
Section~\ref{sec:conclusion} presents our conclusions.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.8\textwidth]{f2a.pdf}
		}

		\subfloat{
			\includegraphics[width=0.8\textwidth]{f2b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf HR diagrams of NGC\,2516, using Gaia EDR3 photometry.}
    {\it Top:} The core (black) shows a clean sequence consistent with
    stars with a fixed age and metallicity, and varying mass.  The
    halo (blue) is similar, but somewhat noisier.  The faintest M
    dwarfs in the core and halo are brighter than in the field star
    comparison sample (gray), consistent with these stars having not
    yet reached the ZAMS.
    {\it Bottom:} Reported members of the halo, as a function of
    galactic latitude. Can the additional scatter in the halo be
    understood through differential reddening?
    {\bf Maybe}.
    \label{fig:hr}
  }
\end{figure*}

@@

%TODO 
Comparing these HR diagrams to the PARSEC isochrone models, we find
that X, Y, and Z.
In particular, ``the faintest M dwarfs in the core and halo are
brighter than in the field star comparison sample, consistent with
these stars having not yet reached the ZAMS.''
This is consistent with the 
main-sequence turn-off being at $Bp-Rp\approx0.05$, which implies an
age of XXX.
The resulting photometric age we adopt for the core is XXX.
For the halo, the claimed age from photometry is YYY.
Applying the same procedure to the field star comparison sample,
we get an age of ZZZ.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.8\textwidth]{f3a.pdf}
		}

		\subfloat{
			\includegraphics[width=0.8\textwidth]{f3b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf The core and halo of \cn in the space of rotation
    period and Gaia color.}
    The {\it top} plot shows periods against a linear scale, while the
    {\it bottom}  shows them against a logarithmic scale.
		\label{fig:rot}
	}
\end{figure*}

@@

\subsection{Lithium from Gaia-ESO and GALAH}
\label{subsec:spectra}

@@

  \subsection{How did it form?}
  \subsection{Mass differences between center and outer reaches?}
  \subsection{Fast rotators: are we going faster than other clusters?}

@@

>>>>

%TODO:
%
% n rotation period logs: 2238 % ~/proj/cdips/results/allvariability_reports/NGC_2516/logs
% n all variability reports: 2201 % ~/proj/cdips/results/allvariability_reports/NGC_2516/reports

@@

%% Reintroduced the \received and \accepted commands from AASTeX v5.2.
%% Add "Submitted to " argument.
\received{---}
\revised{---}
\accepted{---}
\submitjournal{ApJL.}
\shorttitle{Core \& Halo of NGC\,2516}

@@

% Context, aims, methods, results, conclusions.
\begin{abstract}
  Recent analyses of the Gaia data have reported the existence of
  diffuse stellar populations (``halos'') surrounding nearby open clusters.  The stars in these
  halos could have escaped from the cores of the cluster, or they
  could be remnants of a larger and lower-density star formation
  complex that has since dispersed.  They could also be field
  stars---false positives given by overly aggressive clustering algorithms.
  In this study, we focus on a single reported 250\,pc-long halo
  around the $\approx$120\,Myr open cluster NGC\,2516.  The classical tidal radius of
  this cluster is $\sim$XX\,pc.  Combining photometry from Gaia,
  rotation periods from TESS, and lithium measurements from
  the Gaia-ESO and GALAH surveys, we find that: i) the halo of
  NGC\,2516 is real and coeval with the core; ii) at fixed stellar
  mass and age, rapidly rotating stars have greater lithium abundances
  than slowly rotating stars; and iii) given Gaia-selected halo
  members, the contamination rate of field stars in the halo (XX\%) is
  larger than that in the core (YY\%).  This work expands the set of
  confirmed NGC\,2516 members by a factor of $\approx$2, and
  quantifies the degree to which Gaia-based analyses can currently
  identify dispersed stellar populations against a background of field
  star contaminants.  Implications for spectroscopic survey targeting, open cluster dispersal, and planet searches around young
  stars are discussed.
\end{abstract}

@@

In the traditional picture of star formation molecular clouds go
gravitationally unstable, and collapse into little knots.  The knots
produce groups of stars close to each other.  The resulting ``open star
clusters'' are, as with any first-order guess, spherical.

How does the cluster evolve?  The answer depends
on the total stellar mass.
% What exactly does this mean?
The smallest 90\% of star clusters disperse during the ``embedded
phase'' (CITE).  This is driven mostly by {\it process A} (CITE).  The
more massive clusters make it maybe 100 Myr. Maybe a bit longer.
Their evaporation is thought to be driven by collisions with molecular
clouds (CITE, Spitzer 1958), and the galactic tide (CITE), and {\it
process A}, and {\it process B}.  Of course, the expected evaporation
time should depend on factors including the mass of the cluster
itself, and how many high-mass (O and B) stars form, since their winds
[and maybe supernovae] clear out most of the cloud, and the initial
density of the gas cloud to begin with.

@@

We've recently been making TESS light curves of age-dated stars across
the sky, as part of a Cluster Difference Imaging Photometry Survey
(CDIPS, CITE Bouma+19).  Our analysis of Cycle 1 (Sectors 1-13)
yielded light curves of \numsouthernuniqlcs candidate cluster members
in the Southern Ecliptic hemisphere, available on MAST\footnote{\url{https://archive.stsci.edu/hlsp/cdips}}.
% TODO: make this number actually quantitative...
Based on rotation periods, $\approx$25\% appear to be bonafide cluster
members.

As part of a broader project of identifying a large and clean sample
of young stars for a transit search, we focus in his
paper on a rather modest question: in just a single rich
southern open cluster, is the cluster halo coeval with the core?  The
cluster we chose for this analysis was NGC\,2516, since it was young
($\sim$112\,Myr) and close ($d=400$\,pc) enough to facilitate rotation
measurements using TESS, and some degree of spectroscopic analysis.
We want to know: is the halo real? To what extent can we use Gaia
alone to reliably identify age-dated needles in the haystack of boring
field stars?  And more generally, what are the implications for the
evolution of open clusters if they do have halos?

Section~\ref{sec:gaia} presents the astrometric and photometric data
from Gaia, and clarifies our usage of the terms ``core'' and ``halo''.
Section~\ref{sec:agedate} age-dates the halo of NGC\,2516,
based on Gaia photometry (Section~\ref{subsec:hr}),
TESS gyrochronology (Section~\ref{subsec:tess}), and lithium depletion
(Section~\ref{subsec:lithium}).  In Section~\ref{sec:discussion} we
discuss the implications of this analysis for NGC\,2516 specifically
and stellar spin-down and open cluster evolution generally.
Section~\ref{sec:conclusion} presents our conclusions.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.75\textwidth]{f2a.pdf}
		}

		\subfloat{
			\includegraphics[width=0.75\textwidth]{f2b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf HR diagrams of NGC\,2516, using Gaia EDR3 photometry.}
    {\it Top:} The core (black) shows a clean sequence consistent with
    stars with a fixed age and metallicity, and varying mass.  The
    halo (blue) is similar, but somewhat noisier.  The faintest M
    dwarfs in the core and halo are brighter than in the field star
    comparison sample (gray), consistent with these stars having not
    yet reached the ZAMS.
    {\it Bottom:} Reported members of the halo, as a function of
    galactic latitude. Can the additional scatter in the halo be
    understood through differential reddening?
    {\bf Maybe}.
    \label{fig:hr}
  }
\end{figure*}

@@

%TODO 
Comparing these HR diagrams to the PARSEC isochrone models, we find
that X, Y, and Z.
In particular, ``the faintest M dwarfs in the core and halo are
brighter than in the field star comparison sample, consistent with
these stars having not yet reached the ZAMS.''
This is consistent with the 
main-sequence turn-off being at $Bp-Rp\approx0.05$, which implies an
age of XXX.
The resulting photometric age we calculate for the core is XXX.
For the halo, the claimed age from photometry is YYY.
Applying the same procedure to the field star comparison sample,
we get an age of ZZZ.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.75\textwidth]{f3a.pdf}
		}

		\subfloat{
			\includegraphics[width=0.75\textwidth]{f3b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf The core and halo of \cn in the space of rotation
    period and Gaia color.}
    The {\it top} plot shows periods against a linear scale, while the
    {\it bottom}  shows them against a logarithmic scale.
		\label{fig:rot}
	}
\end{figure*}

@@

First, we collected the XXXX Gaia DR2 source\_ids corresponding to the
core and halo members.  For each source, we first retrieved all
available CDIPS light curves, on a per-sector basis.  Then, we
detrended the systematics in each sector individually, and stitched
together the resulting light curves before searching for the
periodicity.  Some details regarding our detrending approach are
discussed in Appendix~\ref{app:detrending}.  After applying a
detrending step aimed at removing systematic trends, we proceeding
with a few small cleaning steps aimed at improving the purity of the
rotation period measurements: we masked 0.7 days at the beginning and
end of each spacecraft orbit, and ran a sliding standard-deviation
rejection window over the light curve, which removed any outlying
points within $\pm3\times$MAD of the median in each window.

We then measured the rotation period of the resulting light curve.  We
used the the aperture radius that, based on theoretical expectations,
was expected to give the optimal balance between light from the target
and background-light (CITE Sullivan15).  This typically resulted in an
aperture radius of either 1 or 1.5 pixels.  To measure the periods, we
used the periodogram implementations in \texttt{astrobase}, in
particular the Stellingwerf PDM periodogram (CITE), along with the
more traditional Lomb-Scargle (CITE).  We recorded the top five
periodogram peaks from each method, and their corresponding powers.
Finally, as a check on crowding, we also recorded the number of stars
within the aperture of equal brightness, and of brightness with 1.25
and 2.5 TESS magnitudes of the target star.

Figure~\ref{fig:rot} shows the resulting rotation periods.  The points
on this plot are light curves for which the peak Lomb Scargle
periodogram period was below 15 days, had power exceeding 0.08, and
had at most one equal-brightness companion within the aperture.  These
selection criteria are entirely heuristic.  
But splitting the sample into the ``core'' and the ``halo'', they
suggest that the halo is real.

The resulting gyrochronology age we find for the core is XXX.
For the halo, the claimed age from gyrochronology is YYY.
Applying the same procedure to the field star comparison sample,
we get an age of ZZZ.

\subsubsection{Kinematics $\otimes$ Rotation}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.95\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Gaia-based components of NGC\,2516 in position and
    velocity space, cross-matched against the rotators.}
		\label{fig:gaia6d_x_rotn}
	}
\end{figure*}

How far away from the core, in position and velocity space, does the
halo actually extend?  We can explore this by crossmatching the
rotator sample against the Gaia data.  Figure~\ref{fig:gaia6d_x_rotn}
shows the result.

\subsection{Lithium from Gaia-ESO and GALAH}
\label{subsec:lithium}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.75\textwidth]{f5a.pdf}
		}

		\subfloat{
			\includegraphics[width=0.75\textwidth]{f5b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Lithium equivalent widths in NGC\,2516,}
  plotted against extinction-corrected color ({\it top}), and the
  combination of rotation period and extinction-corrected color ({\it
  bottom}).
  \label{fig:lithium}
  }
\end{figure*}

Figure~\ref{fig:lithium} shows the results.  At fixed stellar mass
(and age), it seems that the rapid rotators tend to show elevated
lithium equivalent widths.  Similar trends have been previously noted
in the Pleiades by \citet{soderblom_evolution_1993} and
\citet{bouvier_pleiades_lirot_2018}, in the Psc-Eri stream by
\citet{arancibia_2020}, and in M\,35 by \citet{jeffries_m35_li_2020}.

% TODO: this paragraph more or less parrots the processes described by
% Bouvier+20. are there others? is there a better way to describe
% these?
The trend contradicts what might be the naive expectation that rapidly
rotating stars should show more vigorous convection, leading to
earlier destruction of photospheric Li.  Possible physical
explanations include processes both internal and external to the star
(CITE see the recent review by Bouvier+20).  Internal processes would
be tied to the interplay between surface rotation, differential
rotation in the interior, and the convective mixing efficiency (e.g.,
CITE Siess + Livio 1997, Baraffe+2017).  Internal processes could also
include the possibility that stronger magnetic fields in the stellar
interior inhibit convection (READ e.g., Ventura+98, Chabrier+07,
Somers + Pinsonneault 2014).  An external process that could also be
important is the effect of star-disk magnetic braking during the PMS
phase (CITE: magnetic braking).  Longer disk lifetimes would lead to
the star's outer convective zone being ``locked'' for longer while the
radiative core continues contracting.  The resulting differential
rotation and rotational mixing could drive the lithium depletion
(CITE: Bouvier 08, Eggenberger+12).

@@

  \subsection{What is the contamination fraction in the halo? Does it change vs. location?}
  \subsection{Are the ``very slow rotators'' bonafide members?}
  Probably not. They are not isotropically distributed around the
  cluster.
  \subsection{How did the halo form?}
  \subsection{Mass differences between center and outer reaches?}
  \subsection{Fast rotators: are we going faster than other clusters?}

@@

\section{Detrending details}
\label{app:detrending}

In ``detrending'' for our general variability search, our goal was to
preserve astrophysical variability, while removing systematic
variability.  One particular concern for the TESS light curves is
systematic variability at the timescale of the 14-day satellite orbit,
mostly induced by scattered light from the Earth and Moon.

We therefore turned to the principal components (i.e., the
eigenvectors) calculated following the procedure described by
\citet{bouma_cdipsI_2019}. In brief, these vectors are computed using
a set of ``trend stars'' selected from across each CCD according to
ad-hoc heuristics that (hopefully) lead them to be dominated by {\it
systematic} variability (Sec~3.7.2).

The principal component vectors, also referred to as the eigenvectors,
are rank-ordered by the degree of variance that they predict in the
training set (of ``trend stars'').

We then posit that any given target star's light curve is described as
a linear combination of the eigenvectors.  Optionally, we also
considered the inclusion of additional systematic vectors that could
affect the light curve, such as the CCD temperature, the flux level
measured in the background annulus, and the centroid positions of the
stars on the CCDs.  These can be treated as additional ``features'' in
the linear model.

There are many methods for determining the coefficients of the linear
model, after the full set of eigenvectors (plus optionally
``sytematic'' vectors) has been asssembled.  We explored two: ordinary
least squares, and ridge regression. Ridge regression is the same as
ordinary least squares, except it includes an L2 norm with a regularization
coefficient. The regularization coefficient that best applied for any
given target light curve was solved for using a cross-validation grid
search, using \texttt{sklearn.linear\_modelRidgeCV} (CITE).

Each target light curve was mean-subtracted and normalized by its
standard deviation, as were the eigenvectors. The linear problem was
then solved, and the light curve was reconstructed by re-adding the
original mean, and re-multiplying by the standard deviation to ensure
that the variance of the light curve did not change.

We found, somewhat to our surprise, that the choice of using ordinary
least squares versus ridge regression did not seem to significantly
affect the resulting light curves. In other words, the inclusion (or
lack thereof) of a regularization did not strongly alter the
best-fitting coefficients.

A few other choices seemed to be more important:

\begin{itemize}

  \item {\it To smooth, or to not smooth the eigenvectors}.
  Ideally, the eigenvectors should be smooth (and not contain
    residuals from {\it e.g.}, eclipsing binaries that snuck their way
    into the template set). However this does not always apply.
    Sometimes, the eigenvectors are also noisy, which leads them to
    induce extra variability into the PCA ``detrended'' light curves.
    To address this problem, we opted to smooth the eigenvectors using
    a windowed filter (with a "biweight" weight scheme, implemented in
    \texttt{wotan} by \citet{hippke_wotan_2019}; window length 1 day, cval 6).
  One issue with this is that systematic sharp features (captured e.g., in
  "spike vectors") no longer are captured, so they end up in the "PCA
  detrended" light curves. They can be filtered out relatively easily though
  (using rolling outlier rejection), and we prefer this approach to having
  systematic features {\it injected by} the PCA detrending.

  \item {\it How many eigenvectors to use}.
    A larger number always leads to greater whitening.  In
    \citet{bouma_cdipsI_2019}, we performed a Factor Analysis
    cross-validation to determine the number of eigenvectors to use.
    The typical number adopted based on this analysis was 10--15.
    Even though this approach was chosen to prevent over-fitting, in
    our experience, for stellar rotation it still often lead to
    overfitting, especially for rotation signals with periodicities of
    $\gtrsim 3$ days.  (Shorter signals typically are not distorted,
    since the eigenvectors generally do not contain the high-frequency
    content that leads to the distortions).  For this analysis, we
    therefore impose the maximum number of eigenvectors to be 5.

  \item {\it Which supplementary systematics vectors to use}.
    We considered using the \texttt{BGV}, \texttt{CCDTEMP},
    \texttt{XIC}, \texttt{YIC}, and \texttt{BGV} vectors, packaged
    with the CDIPS light curves. we found that the background value
    measured in an annulus centered on the aperture, \texttt{BGV},
    tended to produce the best independent information from the PCA
    eigenvectors, and so we adopted it as our only ``supplementary''
    trend vector.  We opted to not smooth it (in hopes that it would
    provide direct complement to the smoothed PCA vectors; 1 sharp
    vector containing literally the background information, plus 5
    smooth vectors).
\end{itemize}

For every ``target star'', we then decorrelated the raw
(image-subtracted and background-subtracted) light curve
using a linear model with ordinary least squares.

\section{Rotation $\otimes$ RUWE}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.95\textwidth]{f6.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Rotation versus color, colored by RUWE.}
		Looks like on the slow sequence, there's more yellow at the bottom?
		\label{fig:rotn_X_RUWE}
	}
\end{figure*}
